Manuscript draft

### 2019-07-29 manuscript/manuscript draft.Rmd
Update manuscript draft.Rmd

*Model specification.* The effects of climate and latitude on primary productivity were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r:_2018]. The effect of each extracted climate variable on each measure of primary productivity was modelled by specifying the climate variable as a fixed effect. For each climate variable, three models were specified: a null model; a model with the climate variable as a linear term; and a model with the climate variable as a polynomial term. AIC values were calculated for the models and used to select the best model. R-squared values were calculated for the best model. All R-squared values presented here are marginal r-squared values, and refer to the proportion of variation explained by only the fixed effects, unless otherwise specified. In addition, slope coefficients were calculated for the linear models.

@@

To investigate the potential interactive effects of climate variables on carbon fluxes, multivariate models were also specified. Climate variables which were non-significant or poor predictors in the models with a single climate term, or which were strongly correlated with another climate term, were dropped, leaving a subset of 7 climate variables: mean annual temperature; temperature seasonality; mean annual precipitation; precipitation seasonality; potential evapotranspiration; vapour pressure deficit; and solar radiation. For each possible pairing of climate variables, two models were specified: a model with the two climate variables showing an additive effect; and a model with the two climate variables showing an interactive effect. In addition a null model, and models including each climate term as a individual (univariate) linear term were specified. As before, altitude was included as a fixed effect, and plot nested within geographic area as a random effect. AIC values were calculated for the models and used to select the best model. R-squared values were calculated for the best model. Slope coefficients were calculated for each of the terms, and for the interactive term if relevant.

@@

*How do carbon fluxes vary with latitude?* All major carbon fluxes increase with decreasing latitude (fig. 2). Models showed significant variation in whether they were improved by inclusion of altitude as a fixed effect. Including altitude had a very small effect on most models, with the difference in AIC values between models including and excluding altitude often being <2, suggesting the models are very similar in their explanatory power. All models presented below do not include altitude, and r-squared values refer to the proportion of variation explained by the climate variable alone.

@@

Latitude is a strong predictor for many of the carbon fluxes, explaining 66% of variation in GPP (n = 254, p<0.0001), 51% in NPP (n = 116, p<0.0001) and 50% in ANPP (n = 259, p<0.0001). For all fluxes, their relationship with latitude was best predicted by the linear model.

@@

**Mean annual temperature.** All carbon fluxes show a linear increase with temperature.  
**Temperature seasonality and annual temperature range.** These two climate variables are strongly correlated, and so only temperature seasonality is presented in figure 4. Productivity is highest where temperature seasonality = 0, and at an annual temperature range of 15$^\circ$C or lower. GPP, NPP, ANPP, and R~root~ show a polynomial relationship with both variables, and BNPP~root~ shows a polynomial relationship with annual temperature range. As seasonality increases, these variables show a more rapid decrease in productivity. ANPP~foliage~ and ANPP~woody stem~ show a linear relationship with both variables.  
**Mean annual precipitation.** Despite being a poor predictor of productivity, the results indicate that, for the majority of fluxes, productivity increases with precipitation, up until a saturation point at between 3000 and 4000mm annual precipitation, above which productivity starts to decrease. The notable exception to this is GPP, which shows a significant (p<0.0001) linear increase with precipitation up to measures of nearly 6000mm annually, although the explanatory power of this model was low (r-squared = 0.15).  
**Potential evapotranspiration.** All fluxes, with the exception of BNPP~fine root~ and R~root~ show a strong increase with potential evapotranspiration at values below 1000mm. Between 1200 and 1700mm productivity peaks, and begins to decrease at values above 1800mm.

@@

*Do carbon fluxes vary in their responses to climate, particularly with regards to allocation between fluxes?* Of all the flux pairings tested, the only ones to show a significant relationship with latitude were ANPP:NPP (p<0.01, r^2^ = 0.44) and ANPP:BNPP (p<0.001, r^2^ = 0.33). In both cases, allocation to ANPP increased with latitude, with the lowest allocation to ANPP in the tropical regions. BNPP:NPP decreased with latitude, though this relationship was non-significant.

@@

| Variable | Latitude | Mean annual temperature | Temperature seasonality | Potential evapotranspiration | Multivariate model |
|----------|----------|----------|----------|--------------|----------|
| R~auto~ | 0.64 | 0.83 | 0.54 | 0.63 | NA |
| GPP | 0.66 | 0.61 | 0.75 | 0.40 | 0.73 |
| NPP | 0.51 | 0.49 | 0.52 | 0.34 | 0.57 |
| ANPP | 0.50 | 0.45 | 0.47 | 0.32 | 0.50 |
| ANPP~foliage~ | 0.55 | 0.63 | 0.54 | 0.40 | 0.66 |
| ANPP~woodystem~ | 0.25 | 0.23 | 0.20 | 0.26 | NA |
| BNPP~root~ | 0.31 | 0.26 | 0.31 | 0.32 | 0.38 |
| BNPP~fineroot~ | 0.14 | 0.11 | 0.16 | 0.11 | 0.19 |
| R~root~ | 0.25 | 0.21 | 0.29 | 0.04* | 0.43 |

Table: R-squared values presented for latitude, three climate variables, and the best multivariate model for each carbon flux. Values marked * are not significant.

R-squared values are generally highest in the major fluxes, and decrease in subsidiary fluxes. Of the major fluxes, R~auto~ and GPP are the most strongly explained by latitude and climate. Mean annual temperature explains 83% of variation in R~auto~, while temperature seasonality explains 75% of variation in GPP. The proportion of variation explained by climate and latitude decreases in NPP and ANPP. The climatic variables with the strongest explanatory power explain around 50% of the variation in these fluxes. Of the major fluxes, BNPP~root~ is the least well explained by climate and latitude, with climate explaining at most 32% of variation.

With the exception of ANPP~foliage~, the proportion of variation explained by climate and latitude in subsidiary fluxes is much lower. Climate explains at most 26% of variation in ANPP~woody stem~, 17% in BNPP~fine root~, and 32% in R~root~. In contrast, climate strongly explains variation in ANPP~foliage~, with mean annual temperature explaining 63% of variation.

This pattern is also seen in the r-squared values for multivariate models.

>>>>

*Model specification.* The effects of climate and latitude on primary productivity were analysed using mixed effects models using the package 'lme4' [@bates_fitting_2015] in R v.3.5.1 [@r_core_team_r:_2018]. The effect of each extracted climate variable on each measure of primary productivity was modelled by specifying the climate variable as a fixed effect. For each climate variable, three models were specified: a null model; a model with the climate variable as a linear term; and a model with the climate variable as a polynomial term. AIC values were calculated for the models and used to select the best model. R^2^ values were calculated for the best model. All R^2^ values presented here are marginal R^2^ values, and refer to the proportion of variation explained by only the fixed effects, unless otherwise specified. In addition, slope coefficients were calculated for the linear models.

@@

To investigate the potential interactive effects of climate variables on carbon fluxes, multivariate models were also specified. Climate variables which were non-significant or poor predictors in the models with a single climate term, or which were strongly correlated with another climate term, were dropped, leaving a subset of 7 climate variables: mean annual temperature; temperature seasonality; mean annual precipitation; precipitation seasonality; potential evapotranspiration; vapour pressure deficit; and solar radiation. For each possible pairing of climate variables, two models were specified: a model with the two climate variables showing an additive effect; and a model with the two climate variables showing an interactive effect. In addition a null model, and models including each climate term as a individual (univariate) linear term were specified. As before, altitude was included as a fixed effect, and plot nested within geographic area as a random effect. AIC values were calculated for the models and used to select the best model. R^2^ values were calculated for the best model. Slope coefficients were calculated for each of the terms, and for the interactive term if relevant.

@@

*How do carbon fluxes vary with latitude?* All major carbon fluxes increase with decreasing latitude (fig. 2). Models showed significant variation in whether they were improved by inclusion of altitude as a fixed effect. Including altitude had a very small effect on most models, with the difference in AIC values between models including and excluding altitude often being <2, suggesting the models are very similar in their explanatory power. All models presented below do not include altitude, and R^2^ values refer to the proportion of variation explained by the climate variable alone.

@@

Latitude is a strong predictor for many of the carbon fluxes, explaining 64% of variation in GPP (n = 254, p<0.0001), 50% in NPP (n = 114, p<0.0001) and 45% in ANPP (n = 259, p<0.0001). For all fluxes, their relationship with latitude was best predicted by the linear model.

@@

**Mean annual temperature.** All carbon fluxes show a linear increase with temperature.  
**Temperature seasonality and annual temperature range.** These two climate variables are strongly correlated, and so only temperature seasonality is presented in figure 4. Productivity is highest where temperature seasonality = 0, and at an annual temperature range of 15$^\circ$C or lower. GPP, NPP, ANPP, and R~root~ show a polynomial relationship with both variables, and BNPP~root~ shows a polynomial relationship with annual temperature range. As seasonality increases, these variables show a more rapid decrease in productivity. ANPP~foliage~ and ANPP~woody stem~ show a linear relationship with both variables.  
**Mean annual precipitation.** Despite being a poor predictor of productivity, the results indicate that, for the majority of fluxes, productivity increases with precipitation, up until a saturation point at between 3000 and 4000mm annual precipitation, above which productivity starts to decrease. The notable exception to this is GPP: the model indicates that GPP continues to increase with precipitation up to measures of at least 5000mm annually (p<0.0001, R^2^ = 0.33. Data above this point is not available, but the model trend indicates that the saturation point for this model is around 5000mm mean annual precipitation.  
**Potential evapotranspiration.** All fluxes, with the exception of BNPP~fine root~ and R~root~ show a strong increase with potential evapotranspiration at values below 1000mm. Between 1200 and 1700mm productivity peaks, and begins to decrease at values above 1800mm.

@@

*Do carbon fluxes vary in their responses to climate, particularly with regards to allocation between fluxes?* Of all the flux pairings tested, the only ones to show a significant relationship with latitude were ANPP:NPP (p<0.01, R^2^ = 0.44) and ANPP:BNPP (p<0.001, R^2^ = 0.33). In both cases, allocation to ANPP increased with latitude, with the lowest allocation to ANPP in the tropical regions. BNPP:NPP decreased with latitude, though this relationship was non-significant.

@@

| Variable | Latitude | Mean annual temperature | Temperature seasonality | Potential evapotranspiration | Multivariate model |
|----------|----------|----------|----------|--------------|----------|
| R~auto~ | 0.65 | 0.77 | 0.62 | 0.63 | NA |
| GPP | 0.64 | 0.63 | 0.72 | 0.36 | 0.73 |
| NPP | 0.50 | 0.52 | 0.52 | 0.33 | 0.57 |
| ANPP | 0.45 | 0.46 | 0.42 | 0.28 | 0.50 |
| ANPP~foliage~ | 0.54 | 0.63 | 0.54 | 0.35 | 0.66 |
| ANPP~woodystem~ | 0.17 | 0.23 | 0.12 | 0.19 | NA |
| BNPP~root~ | 0.33 | 0.26 | 0.23 | 0.36 | 0.38 |
| BNPP~fineroot~ | 0.15 | 0.11 | 0.16 | 0.11 | 0.19 |
| R~root~ | 0.22 | 0.21 | 0.30 | 0.01* | 0.43 |

Table: R^2^ values presented for latitude, three climate variables, and the best multivariate model for each carbon flux. Values marked * are not significant.

R^2^ values are generally highest in the major fluxes, and decrease in subsidiary fluxes. Of the major fluxes, R~auto~ and GPP are the most strongly explained by latitude and climate. Mean annual temperature explains 77% of variation in R~auto~, while temperature seasonality explains 72% of variation in GPP. The proportion of variation explained by climate and latitude decreases in NPP and ANPP. The climatic variables with the strongest explanatory power explain around 50% of the variation in these fluxes. Of the major fluxes, BNPP~root~ is the least well explained by climate and latitude, with climate explaining at most 36% of variation.

With the exception of ANPP~foliage~, the proportion of variation explained by climate and latitude in subsidiary fluxes is much lower. Climate explains at most 23% of variation in ANPP~woody stem~, 16% in BNPP~fine root~, and 30% in R~root~. In contrast, climate strongly explains variation in ANPP~foliage~, with mean annual temperature explaining 63% of variation.

This pattern is also seen in the R^2^ values for multivariate models.
